---
title: "Assignment 2: Coding Basics"
author: "Eva May"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your first and last name into the file name (e.g., "FirstLast_A02_CodingBasics.Rmd") prior to submission.



## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}
#1. cent <- seq(1, 100, 4) #cent is name of sequence
#(from, to, by)

#2. mean(cent); median(cent)
#mean of created sequence and median of created sequence
#; separates 2 commands like separate lines of code would

#3. mean(cent) > median(cent)
#using > returns T/F statement about the computed numbers from the 2 commands

```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}
#5 and 6

name <- c("Anna", "Jeff", "Zoe", "Bob")
typeof(name)
#character vector - contains a list of names

score <- c(65, 40, 80, 70)
typeof(score)
#double numeric vector - automatic kind of vector created for integers is a double-precision numerical one

passed <- function(x){
  ifelse(x>49, TRUE, FALSE)
}
#created a function with ifelse statement to describe if the test scores in score passed - making this vector correspond with vector score
pass <- c(passed(score))
typeof(pass)
#logical vector - contains logical values of F and T


#7

student.scores <- data.frame(name, score, pass)
#creating data frame - each vector becomes a column

#8

colnames(student.scores)
#getting current column names
names(student.scores)[names(student.scores) == "name"
                      ] <- "Student Name"
names(student.scores)[names(student.scores) == "score"
                      ] <- "Test Score"
names(student.scores)[names(student.scores) == "pass"
                      ] <- "Student Passed?"
#changing each column name individually

student.scores
#returning data frame to make sure names changed
```

9. QUESTION: How is this data frame different from a matrix?

> Answer: Each column in this data frame contains a different type of data - character, numeric, and logical. A matrix would not allow this, and would require that all data be the same type (e.g. 3 character columns).

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative. 

>Answer [oops sorry - kind of did this above so will just adapt that one]

```{r}
passing.score <- function(x){
  ifelse(x>49, TRUE, FALSE)
}
#returns true if x>49 and false if x<=49
#did not include 'print' here as function automatically returns either TRUE or FALSE. including 'print' is redundant and causes function to print TRUE and FALSE in addition to the logic responses related to each x. 
```


11. Apply your function to the vector with test scores that you created in number 5. 
> [again sorry I did this above..hope that's okay!]

```{r}

passing.score(score)
#applying fxn to vector of test scores

passpass <- function(x){
  if(x>49) TRUE
  else FALSE
}
#creating if and else fxn to see if it will also work
passpass(score)
#error because score is a vector with length >1
passpass(2)
#no error bc only 1 numerical value is input
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: I used ifelse rather than the longer if and else, but I initially thought both would work because they do the same thing - I just chose to use the quicker one. However, in the block of code above, I tried creating an if and else function as well (passpass), which R accepts as a function but won't run on the vector, as it seems to only work with single numerical values, e.g. scalars (and ifelse are designed more for vectors with length > 1). So, I would say that the ifelse function works for this scenario because we are running it on a vector. You'll also see in the code above that if passpass is run on a single numerical value, R is fine with it, so if score were a vector with only 1 value, either function would work.
